Hydras and Cie (Draft) \\ Variations on Kirby \& Paris' hydra battles and ordinal numbers in Coq

### 2021-09-17 doc/JFLA2022/paper.tex
minor corrections

\newtheorem{todo}{To do}
\usepackage{amsfonts, afterpage}

@@

\begin{abstract}
  \begin{todo}
 \dots
  \end{todo}
\end{abstract}

@@

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}.
This article describes a game between two players: Hercules and a hydra.
A short description  can be found in Andrej Bauer's blogpost~\cite{bauer2008} as well as a simulator~\cite{BauerHydra}.

In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionnaly presented with the root at the bottom, the leaves of which are called \emph{heads}.
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $i$ of times. The number $i$ of copies and the copied sub-hydra $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
\end{itemize}

@@

The contrast between the simplicity of the statements above and the complexity of the mathematical tools used to prove these theorems: ordinal numbers, rapidly growing hierarchies of fast growing functions, etc., convinced us that it is a good thema for a commented library of formal proofs.

The benefits we would expect from this development are multiple.
\begin{itemize}
\item We present complete proof scripts which can be consulted and replayed by the reader.
\item Complex formalisations and proofs are explained in an
  electronic book~\cite{HydraBook}. Whenever various reasonable choices exist, we try to present and compare the alternatives. Reciprocally, we provide ``concrete'' examples of several formaisation end proof techniques: operational type classes, functions defined by  equations, etc.
\item It may be used as a library on ordinal numbers, for instance for proving termination properties.
  \end{itemize}

 Prior stages of this project have already been presented at
 JFLA~\cite{JFLA2018paper}.
We present the recent evolutions of the library: new results, interaction within the Coq-Community project~\cite{CoqCommunity}, and documentation generated with Alectryon~\cite{alectryonpaper, alectryongithub}.

@@

   \item The hydra grows $n$ copies at step $n$ of the game.
 \item  The initial hydra is the graphical representation of an ordinal  $\alpha<\epsilon_0$  in Cantor normal form (See for instance Fig~\ref{fig:iota-example}). 
 \item Hercules always chops off the rightmost head of the hydra (which is always one of the lowest heads).
 \end{itemize}
 It mathematical terms, if at step $n$ the hydra is associated with the ordinal $\alpha$, at step $n+1$ it is associated with
 $\canonseq{\alpha}{n+1}$, the $(n+1)$th element of the canonical sequence of $\alpha$~\cite{KS81}.

@@

Let $\alpha<\epsilon_0$ be an ordinal.
Using the \texttt{coq-equations} plug-in~\cite{sozeau:hal-01671777}, we define the function which returns the number of steps of the battle starting at configuration $(i,\alpha)$ for  any $i\in\mathbb{N}$.
We prove that this number is greater or equal than
$H'_\alpha(i)-i$, where $H'$ is a slight variant of the Wainer Hardy hierarchy of rapidly growing functions~\cite{BW85, KS81, Promel2013, Wainer1970}, defined in Fig.~\ref{fig:Hprime} and
\ref{fig:HprimeCoq}.

\begin{figure}[h]
  \centering
  \begin{itemize}
\item If $\alpha=0$, then $H'_\alpha (k)= k$ for any natural number $k$.
\item If $\alpha=\beta+1$, then 
$H'_\alpha(k)=H'_\beta(k+1)$ for any $k \in \mathbb{N}$
\item If $\alpha$ is a limit ordinal, then 
$H'_\alpha(k) = H'_{(\canonseq{\alpha}{k+1})}(k)$ for any $k\in \mathbb{N}$.
\end{itemize}
\caption{The rapidly growing hierarchy $H'$}
\label{Hprime}
\end{figure}

\begin{figure}[h]
  \centering
  \inputsnippets{Hprime/HprimeDef}
  \caption{$H'$: definition in \emph{Coq}}
  \label{fig:HprimeCoq}
\end{figure}

From this result, we infer that, for a given $\alpha$, the considered function is not primitive recursive.

@@

% \begin{eqnarray}
%   \label{eq:1}
%   \omega+42+\omega^2&=&\omega^2
% \end{eqnarray}

% We also implemented in \emph{Coq} the axiomatic definition of the set $\mathbb{O}$ of countable ordinals by Schütte~\cite{schutte, PCiota}.

% The proof of $\omega+42+\omega^2=\omega^2$ is no more computational, but use the following definitions and properties.

% \begin{itemize}
% \item Any set of countable ordinals can be enumerated by a function defined on a segment $[0, \alpha)$ of countable ordinals.
% \item Let $\alpha$ and $\beta$ be (countable) ordinals. Then $\alpha+\beta$ is the $\beta$-th element of the enumeration of the set of countable ordinals greater or equal than $\alpha$.
% \item Let $\phi_0$ be the enumeration of the set of \emph{additive principal ordinals}
%   $\{\alpha\in\mathbb{O}| \forall \beta\in \mathbb{O}, \beta+\alpha=\alpha\}$. Then $\omega^\alpha$ is just a notation for
%   $\phi_0(\alpha)$.
% \end{itemize}

\begin{todo}
  Comment the multiple views of a same mathematical object:
  \begin{itemize}

  \item Bridge lemmas (relative correctness of different representations).
    \item The user/reader may compare  different proofs of the same result (For instance the equality $\omega+42+\omega^2= \omega^2$). \emph{Example in Figures~\ref{fig:Ex42E0} and \ref{fig:Ex42-schutte}}.
  \end{itemize}
\end{todo}

\section{The Coq-community ecosystem}

\section{Documentation with Alectryon}

We present two ~examples of the documentation generated with Alectryon~\cite{alectryonpaper}: A simple proof by computation (Fig.~\ref{fig:Ex42E0}), and a longer proof where only the main steps are displayed, interleaved with plain text.

@@

\label{fig:Ex42-schutte}
\fbox{\begin{minipage}[h]{1.0\linewidth}
  Let us prove again the equality $\omega+42+\omega^2= \omega^2$. Let us recall that $\omega^2$ is an abbreviation of $\phi_0(2)$,
\emph{i.e} the third  additive principal ordinal.

@@

\end{minipage}
}
\caption{A proof interleaved with text (from the book)}
  \label{fig:Ex42b}
\end{figure}

@@

\begin{todo}
  \begin{itemize}
  \item  Soundness (correction, maintenance) of the Alectryon generated doc.
  \item Communications between co-maintained libraries
          (Ackermann, Gaia, etc.)
  \item Work destined both to Coq-users and people interested in discrete math ?
  \end{itemize}
\end{todo}

@@

It would be interesting to build a library which could be used as a gate giving access to both the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}.
For instance, the associativity of multiplication of ordinals below $\epsilon_0$, already proven in \textit{Gaia} could be tranferred ``for free'' to the data structures of \textit{hydra-battles}, thanks to the isomorphism between both implementations.

>>>>

\usepackage{varioref}
\newtheorem{todo}{To do}
\usepackage{amsfonts, afterpage}

@@

\begin{abstract}
  We present the current state of a \textit{Coq} library on ordinal numbers.

\end{abstract}

@@

%------------------------------------------------------------------------------
\section{Introduction}
\label{sect:introduction}
Hydra games (also known as \emph{Hydra battles}) appear in an article published in 1982 by two mathematicians, 
Laurie Kirby and Jeff Paris~\cite{KP82}: \emph{Accessible Independence Results for Peano Arithmetic}.
This article describes a game between two players: Hercules and a hydra.
A short description  can be found in Andrej Bauer's blogpost~\cite{bauer2008} as well as a simulator~\cite{BauerHydra}.
In a few words:
\begin{itemize}
\item A hydra is a finite tree, traditionnaly presented with the root at the bottom, the leaves of which are called \emph{heads}
  (Fig~\vref{fig:iota-example}).
\item At every round, Hercules chops off one head of the hydra. If the head is at a distance greater or equal than 2 from the root,
  then some sub-tree $h$ of the hydra is copied a certain amount $i$ of times. The number $i$ of copies and the copied sub-hydra $h$ may depend of the considered variant of the game
  or the time elapsed since the beginning of the fight.
\end{itemize}

@@

The contrast between the simplicity of the statements above and the complexity of their proofs convinced us that it is a good thema for a commented library of formal proofs.

\begin{itemize}
\item We present complete proof scripts for the \textit{Coq} proof assistant~\cite{Coq}. These scripts may consulted and replayed by the reader.
\item Complex formalisations and proofs are explained in an
  electronic book~\cite{HydraBook}. Whenever various reasonable choices exist, we try to present and compare the alternatives.
  For instance, Figures~\vref{fig:Ex42E0} and \vref{fig:Ex42-schutte} show two radically different proofs of the equality
  $\omega+42+\omega^2=\omega^2$, the first one is a simple proof by computation, the second one shows how this equality
  is a consequence of the axioms of the set-theoretic model  by Kurt Schütte~\cite{schutte}. 
\item
  Reciprocally, we provide ``concrete'' examples of several formalisation and proof techniques: operational type classes, functions defined by  equations, dependently typed functions, etc.
\item It may be used as a library on ordinal numbers, for instance for proving termination properties.
  \end{itemize}

 Prior stages of this project have already been presented at
 JFLA~\cite{PCiota, JFLA2018paper}.
We present the recent evolutions of the library: new results, interaction within the Coq-Community project~\cite{CoqCommunity}, and documentation generated with Alectryon~\cite{alectryonpaper, alectryongithub}.

@@

   \item The hydra grows $n$ copies at step $n$ of the game.
   \item  The initial hydra is the graphical representation of an ordinal  $\alpha<\epsilon_0$  in Cantor normal form (See for instance Fig~\ref{fig:iota-example}). This property is an invariant of the game.
 \item Hercules always chops off the rightmost head of the hydra (which is always one of the lowest heads).
 \end{itemize}
 It mathematical terms, if at step $n$ the hydra is associated with the ordinal $\alpha$, at step $n+1$ it is associated with
 $\canonseq{\alpha}{n+1}$, the $(n+1)$th element of the canonical sequence of $\alpha$~\cite{KS81}.

@@

Let $\alpha<\epsilon_0$ be an ordinal.
Using the \texttt{coq-equations} plug-in~\cite{sozeau:hal-01671777}, we define the function which returns the number of steps of the battle starting with $\alpha$ at step $i$ for  any $i\in\mathbb{N}$.
We prove that this number is greater or equal than
$H'_\alpha(i)-i$, where $H'$ is a slight variant of the Wainer Hardy hierarchy of rapidly growing functions~\cite{BW85, KS81, Promel2013, Wainer1970}, defined by transfinite recursion in Fig.~\ref{fig:Hprime}.

\begin{figure}[h]
  \centering
  \begin{itemize}
\item If $\alpha=0$, then $H'_\alpha (i)= i$ for any natural number $i$.
\item If $\alpha=\beta+1$, then 
$H'_\alpha(i)=H'_\beta(i+1)$ for any $i \in \mathbb{N}$
\item If $\alpha$ is a limit ordinal, then 
$H'_\alpha(i) = H'_{(\canonseq{\alpha}{i+1})}(k)$ for any $i\in \mathbb{N}$.
\end{itemize}
\vspace{4pt}
\inputsnippets{Hprime/HprimeDef}
\caption{The rapidly growing hierarchy $H'$}
\label{fig:Hprime}
\end{figure}

\label{sect:not-pr}
From this result, we infer that the considered function is not primitive recursive in general (for $\alpha\geq\omega^\omega$).

@@

\subsection{The Coq-community ecosystem}
\begin{todo}
Small presentation.
\end{todo}

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, weused the formalisation of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem \cite{OConnor05}. We chose to host and maintain this work in Coq-community. The pedagogical interest of this formalisation led us to add new examples and exercises on primtive recursive functions.

\subsection{Documentation with Alectryon}
\begin{todo}
  Interest of Alectryon:
  \begin{itemize}
  \item Despite the frequent changes (improvements) in the Coq scripts, code inclusion (and \textit{Coq}'s anwers are up to date.

  \item Presentation of the main steps of proofs, possibly interleaved with text (Figures \vref{fig:Ex42E0} and \vref{fig:Ex42-schutte},).

  \end{itemize}
\end{todo}

@@

\fbox{\begin{minipage}[h]{1.0\linewidth}
  Let us prove again the equality $\omega+42+\omega^2= \omega^2$. Let us recall that $\omega^2$ is an abbreviation of $\phi_0(2)$,
\emph{i.e} the third  additive principal ordinal.

@@

\end{minipage}
}
\caption{A proof interleaved with text (from the book)}
  \label{fig:Ex42-schutte}
\end{figure}

@@

\begin{todo}
  Work destined both to Coq-users and people interested in discrete math ?
 \end{todo}

@@

It would be interesting to build a library which could be used as a gate giving access to both the combinatorial results of \textit{hydra-battles} and the set-theoretic content of \textit{Gaia}.
For instance, the associativity of multiplication of ordinals below $\epsilon_0$, already proven in \textit{Gaia} was tranferred ``for free'' to the data structures of \textit{hydra-battles}, thanks to the isomorphism between both implementations.
